Gravitational-wave sensitivity curves

### 2014-07-01 Detectors.tex
random tiny change for test purposes
completely rewritten sec4.4 on PTAs

\section{Detectors}\label{sec:detectors}
\subsection{Operating principle of an interferometric detector}\label{sec:principles}
All of the man-made detectors discussed in this section work on the principle of interferometry. Such detectors work by taking a beam of monochromatic light and splitting it into two beams travelling at some angle to each other. Each beam is passed in to a Fabry-P\'{e}rot cavity where it undergoes a number of round trips before being recombined to form an interference pattern. The ends of the cavity are, in the ideal case, freely floating test masses which move in response to a passing GW, this effect is measured by observing the changing interference pattern.

The response of a detector to an incident plane fronted GW depends upon the relative orientations of the detector and the incoming wave. Choose the origin of our coordinate system to be the beam-splitter of the interferometer and let $l_{1}^{i}$ and $l_{2}^{i}$ be unit 3-vectors pointing along the two arms. In the absence of noise the output of the detector is the difference in strain between the two arms \citep{MTW}
\begin{equation}\label{eq:hoftxx} h(t)=\frac{1}{2}h_{ij}\left( l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j} \right)\; , \end{equation}
where $h_{ij}$ are the spatial components of the GW metric perturbation. Let $n^{i}$ be the unit 3-vector pointing towards the source of the GWs, with spherical polar angles $(\theta,\phi)$ relative to some axes fixed to the detector, also let $p^{i}$ and $q^{i}$ be unit vectors orthogonal to $n^{i}$. There remains a freedom in the coordinates described, a rotation of $p^{i}$ and $q^{i}$ through an angle $\psi$ about $n^{i}$. Define the basis tensors,
\begin{eqnarray}
H^{+}_{ij}&=p_{i}p_{j}-q_{i}q_{j} \nonumber \\
H^{\times}_{ij}&=p_{i}q_{j}+q_{i}p_{j} \; .
\end{eqnarray}
Consider a single frequency component of the incident GW, the strain induced by the incident GW may be written as
\begin{equation}\label{eq:hijxx} h_{ij}=A_{+}H^{+}_{ij}\cos\left(2\pi ft\right)+A_{\times}H^{\times}_{ij}\cos\left(2\pi ft+\Delta \phi\right) \; ,\end{equation}
where $A_{+}$ and $A_{\times}$ are the amplitudes of the two polarisation states. Combining (\ref{eq:hoftxx}) and (\ref{eq:hijxx}) allows the detector output to be written as
\begin{equation} h(t)=F^{+}(\theta,\phi,\psi)A_{+}\cos\left(2\pi ft\right)+F^{\times}(\theta,\phi,\psi)A_{\times}\cos\left(2\pi f t + \Delta\phi \right)\; , \end{equation}
where the response functions inherit their angular dependence from the choice of coordinates,
\begin{eqnarray}
F^{+}(\theta,\phi,\psi)&=\frac{1}{2}H^{+}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \nonumber \\
F^{\times}(\theta,\phi,\psi)&=\frac{1}{2}H^{\times}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \; . 
\end{eqnarray}
The response function of a two-arm interferometric detector is quadrupolar, an example is plotted in figure \ref{fig:LIGO}.

@@

\subsection{Pulsar timing arrays (PTAs)}\label{sec:PTAgeneralproperties}
\cjm{I am going to rewrite this section and give a similar level of detail as was given for interferometric detectors in section \ref{sec:principles}.}

PTAs can be thought of as naturally occurring interferometers with galactic scale arm lengths, hence they are sensitive to much lower frequencies than the detectors considered so far. Each pulsar is a very regular clock and the measured arrival time can be compared against a prediction leaving a residual which includes the effects of passing GWs. \cite{SesanaVecchioColancino} describe the shape of the sensitivity curve obtained by correlating timing residuals from each of $N_{\textrm{p}}$ pulsars in an array. The total timing residual is the sum of the residuals due to noise and due to GWs,
\begin{equation} \delta t= \delta t_{n} + \delta t _{h} \; .\end{equation}
The maximum timing residual caused by a passing GW is the light travel time of along wavelength times the strain amplitude of the GW, $\delta t_{h}\approx h/f$. An upper limit to any background may be placed by assuming the residuals are due entirely to GWs. First consider the case of two pulsars; the minimum instantaneous detectable timing residual due to GWs is proportional to the root mean square error in the measurement of the timing residual, $\delta t_{h}\approx\delta t_{\textrm{r.m.s}}$. The minimum detectable signal decreases with the number of measured timing residuals (or data points). Over an observation time $T$, with individual observations taken at intervals of $\Delta T$ the minimum detectable signal is given by
\begin{equation}\label{eq:GWtimingresidual} \delta t_{h} \propto \frac{\delta t_{\textrm{r.m.s}}}{\sqrt{T\Delta T ^{-1}}} \; . \end{equation}
The PTA is sensitive to a minimum GW frequency of $T^{-1}$ and a maximum frequency of $\Delta T^{-1}$, hence the bandwidth is $\Delta f \approx \Delta T^{-1}-T^{-1}$. Assuming $\Delta T \ll T$ gives $T\Delta T^{-1} \approx T \Delta f$. Therefore the minimum characteristic strain the two-pulsar PTA is sensitive to is given by
\begin{equation} h_{c}=\frac{\delta t_{\textrm{r.m.s.}}f}{\left(T\Delta f\right)^{1/4}} \; . \end{equation}
When using more than two pulsars each separate pair of pulsars forms an independent detector, so the number of detectors scales as $N_{\textrm{p}}^{2}$. The optimum SNR is given by adding the individual SNRs in quadrature. Hence the total sensitivity is given by
\begin{equation}\label{eq:PTA} h_{c}(f) \propto \frac{\delta t_{\textrm{r.m.s.}}f}{N_{\textrm{p}}^{1/2}\left( T\Delta f \right)^{1/4}} \; . \end{equation}
The characteristic strain that the PTA is sensitive to scales linearly with $f$ between the minimum and maximum frequencies. This gives the wedge shaped curves plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega}. The absolute value of the sensitivity is fixed by normalising (\ref{eq:PTA}) to agree with a calculated limit at a given frequency for each PTA.

There is a discrepancy between the treatment of PTA sensitivity curves here and the higher frequency detectors discussed in sections \ref{sec:ground} and \ref{sec:space}. When observing a long lived source, such as an inspiral, with a high frequency detector the convention was to define a \emph{characteristic} strain to satisfy (\ref{eq:hc}). Here the convention is to leave the strain untouched and instead adjust the PTA sensitivity curve with observation time, again to satisfy (\ref{eq:hc}). This discrepancy is an unfortunate result of the conventions in use by the different GW communities; however it is also natural given the sources under observation. When observing a transient source, such as a burst or inspiral, which changes within the lifetime of the detector it is natural to consider the detector as performing constantly while the signal changes. However when observing a monochromatic source or a stochastic background which is unchanging over the mission lifetime it is more natural to consider the source as being fixed and the sensitivity of the detector gradually improving. All that is required by the definition in (\ref{eq:hc}) is that the ratio $h_{c}(f)/h_{n}(f)$ is constant.

>>>>

\section{Detectors}\label{sec:detectors}

\subsection{Operating principle of an interferometric detector}\label{sec:principles}
All of the man-made detectors discussed in this section work on the principle of interferometry. Such detectors work by taking a beam of monochromatic light and splitting it into two beams travelling at some angle to each other. Each beam is passed in to a Fabry-P\'{e}rot cavity where it undergoes a number of round trips before being recombined to form an interference pattern. The ends of the cavity are, in the ideal case, freely floating test masses which move in response to a passing GW, this effect is measured by observing the changing interference pattern.

The response of a detector to an incident plane fronted GW depends upon the relative orientations of the detector and the incoming wave. Choose the origin of our coordinate system to be the beam-splitter of the interferometer and let $l_{1}^{i}$ and $l_{2}^{i}$ be unit 3-vectors pointing along the two arms. In the absence of noise the output of the detector is the difference in strain between the two arms \citep{MTW}
\begin{equation}\label{eq:hoftxx} h(t)=\frac{1}{2}h_{ij}\left( l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j} \right)\; , \end{equation}
where $h_{ij}$ are the spatial components of the GW metric perturbation. Let $n^{i}$ be the unit 3-vector pointing towards the source of the GWs, with spherical polar angles $(\theta,\phi)$ relative to some axes fixed to the detector, also let $p^{i}$ and $q^{i}$ be unit vectors orthogonal to $n^{i}$. There remains a freedom in the coordinates described, a rotation of $p^{i}$ and $q^{i}$ through an angle $\psi$ about $n^{i}$. Define the basis tensors,
\begin{eqnarray}
H^{+}_{ij}&=p_{i}p_{j}-q_{i}q_{j} \nonumber \\
H^{\times}_{ij}&=p_{i}q_{j}+q_{i}p_{j} \; .
\end{eqnarray}
Consider a single frequency component of the incident GW, the strain induced by the incident GW may be written as
\begin{equation}\label{eq:hijxx} h_{ij}=A_{+}H^{+}_{ij}\cos\left(2\pi ft\right)+A_{\times}H^{\times}_{ij}\cos\left(2\pi ft+\Delta \phi\right) \; ,\end{equation}
where $A_{+}$ and $A_{\times}$ are the amplitudes of the two polarisation states. Combining (\ref{eq:hoftxx}) and (\ref{eq:hijxx}) allows the detector output to be written as
\begin{equation} h(t)=F^{+}(\theta,\phi,\psi)A_{+}\cos\left(2\pi ft\right)+F^{\times}(\theta,\phi,\psi)A_{\times}\cos\left(2\pi f t + \Delta\phi \right)\; , \end{equation}
where the response functions inherit their angular dependence from the choice of coordinates,
\begin{eqnarray}\label{eq:responsefuncs}
F^{+}(\theta,\phi,\psi)&=\frac{1}{2}H^{+}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \nonumber \\
F^{\times}(\theta,\phi,\psi)&=\frac{1}{2}H^{\times}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \; . 
\end{eqnarray}
The response function of a two-arm interferometric detector is quadrupolar, an example is plotted in figure \ref{fig:LIGO}.

@@

\subsection{Pulsar timing arrays (PTAs)}\label{sec:PTAgeneralproperties}
\cjm{
PTAs can be thought of as naturally occurring interferometers with galactic scale arm lengths, hence they are sensitive to much lower frequencies than the detectors considered so far. Each pulsar is a very regular clock and the measured pulse arrival time (in practice, arrival times are corrected to arrival times at the solar-system  barycenter, SSB) can be compared against a prediction leaving a residual which includes the effects of passing GWs. Using an array of these pulsars spread across the sky allows us to correlate residuals between different pulsars and exploit the fact that the GWs influence all pulsars where as intrinsic pulsar noise clearly does not. The correlation between different pulsars depends only on their angular separation on the sky and has a distinctive shape, known as the Hellings and Downs curve \citep{HellingsDowns}, which we will now derive.

The redshift of the rate of arrival of pulses for a pulsar at a distance $L$ from Earth, in the direction of the unit spatial vector $\vec{p}$ induced by a GW travelling in the direction of the unit vector $\hat{\Omega}$ is given by \citep{anholm-2009}
\begin{equation}\label{eq:pulsarredshift} z(t,\hat{\Omega})=\frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\left(h_{ij}^{\textrm{Pulsar}}(t-L,\hat{\Omega} )-h_{ij}^{\textrm{Earth}}(t,\hat{\Omega} )\right)=\frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\Delta h_{ij}(t,\hat{\Omega})\; . \end{equation}
The redshift includes two terms, the pulsar term and the Earth term. The pulsar term is often neglected in PTA analysis as it can be considered as an extra noise term which averages to zero. The experimentally measured quantity is not the redshift but the timing residual, the two are related via
\begin{equation}\label{eq:restored}  R(t,\hat{\Omega})=\int_{0}^{t}\textrm{d}t'\;z(t',\hat{\Omega}) \; . \end{equation}
All of the pulsars, and the Earth, reside in the same metric perturbation field which may be expressed in terms of its Fourier transform
\begin{equation} h_{ij}(t,\hat{x})=\sum_{A=+,\times}\int\textrm{d}f\;\iint_{S_{2}}\textrm{d}\hat{\Omega}\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega})e^{ 2\pi i f (t-\hat{\Omega}\cdot\vec{x}) } \; ,\end{equation}
where $e^{A}_{ij}(\hat{\Omega})$ is the $A$ polarisation basis tensor for direction $\hat{\Omega}$. Choosing the SSB as the origin of our coordinate system, so the pulsar is at position $L\hat{p}$, gives
\begin{equation}\label{eq:deltahij} \Delta h_{ij}(t,\hat{\Omega})=\sum_{A=+,\times}\int\textrm{d}f\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega})e^{ 2\pi i f t }\left(e^{-2\pi\rmi fL (1+\hat{p}\cdot\hat{\Omega})}-1\right) \; . \end{equation}
From (\ref{eq:pulsarredshift}) and (\ref{eq:deltahij}) the Fourier transform of the redshift, $\tilde{z}(f,\hat{\Omega})$, may be read off as
\begin{equation}\label{eq:45} \tilde{z}(f,\hat{\Omega})=\left(e^{-2\pi\rmi fL (1+\hat{p}\cdot\hat{\Omega})}-1\right)\sum_{A=+,\times} \tilde{h}_{A}(f,\hat{\Omega})F^{A}(\hat{\Omega})\;, \textrm{where}\; F^{A}(\hat{\Omega})=\frac{e_{ij}^{A}(\hat{\Omega})\hat{p}^{j}\hat{p}^{i}} {2(1+\hat{\Omega}\cdot\hat{p})} \, .\end{equation}
The function $F^{A}(\hat{\Omega})$ may be regarded as the PTA equivalent of the detector response functions in (\ref{eq:responsefuncs}). The stochastic background of GWs is fully characterised by the one-sided power spectral density via the following expectation value
\begin{equation}\label{eq:stochback} \left<h_{A}^{*}(f,\hat{\Omega})h_{A'}(f',\hat{\Omega}')\right>=\frac{1}{2}S_{h}(f)\delta^{(2)}(\hat{\Omega},\hat{\Omega}')\delta_{AA'}\delta(f-f') \; ,\end{equation}
where $\delta^{(2)}(\hat{\Omega},\hat{\Omega}')$ is the delta function on the sphere. From (\ref{eq:stochback}) and (\ref{eq:45}) the expectation of the product of signals from two different pulsars in directions $\hat{p}_{1}$ and $\hat{p}_{2}$ may be evaluated
\begin{eqnarray} & \left<z_{1}(f)z_{2}^{*}(f')\right>=\frac{1}{2}S_{h}(f)\delta (f-f')\Gamma(f)\quad \textrm{where,} \\
& \Gamma(f)=\sum_{A=+,\times}\iint_{{\cal{S}}^{2}}\textrm{d}\hat{\Omega}\,\left(e^{2\pi\rmi fL_{1}(1+\hat{\Omega}\cdot\hat{p}_{1})}-1\right)\left(e^{-2\pi\rmi fL_{2}(1+\hat{\Omega}\cdot\hat{p}_{2})}-1\right)F_{1}^{A}(\hat{\Omega})F_{2}^{A}(\hat{\Omega}) \; . \nonumber\end{eqnarray}
The overlap function, $\Gamma(f)$, tends to a constant value in the limit that the distances to the pulsars are large compared to the wavelength of GWs, PTAs operate in this limit, so the overlap may be approximated as a constant 
\begin{equation} \Gamma(f)\approx\Gamma_{0}= \sum_{A=+,\times}\iint_{{\cal{S}}^{2}}\textrm{d}\hat{\Omega}\;F_{1}^{A}(\hat{\Omega})F_{2}^{A}(\hat{\Omega})\;.\end{equation}
Neglecting the exponential terms in the overlap is the frequency domain equivalent of neglecting the pulsar term in (\ref{eq:pulsarredshift}). The integral may be evaluated to give an expression depending only on the angle, $\theta$, between the two pulsars; this is the famous Hellings and Downs curve, see figure \ref{fig:HnD},
\begin{equation} \Gamma_{0}=\frac{1}{2}+\frac{3x}{2}\left(\log (x)-\frac{1}{6}\right) \quad \textrm{where}\quad x=\frac{1-\cos\theta}{2}\; .\end{equation}

\begin{figure}[h!]
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.6\textwidth]{HnDcurve.pdf}
 \caption{The Hellings and Downs curve, the correlation between two pulsars seperated on the sky by an angle $\theta$.}
 \label{fig:HnD}
\end{figure}

The sensitivity bandwidth of a PTA is set by the sampling properties of the data set. If measurements are spaced in time by $\delta t$ and taken for a total length of time $T$ then the PTA is sensitive to frequencies in the range $(1/T)<f<(1/\delta t)$. The characteristic strain that the PTA is sensitive to scales linearly with $f$ in this range. This gives the wedge shaped curves plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega}. The absolute value of the sensitivity is fixed by normalising to a calculated limit at a given frequency for each PTA. For a discussion of the sensitivities of PTA to both individual sources and stochastic background see \cite{MooreTaylorGair}.

There is a discrepancy between the treatment of PTA sensitivity curves here and the higher frequency detectors discussed in sections \ref{sec:ground} and \ref{sec:space}. When observing a long lived source, such as an inspiral, with a high frequency detector the convention was to define a \emph{characteristic} strain to satisfy (\ref{eq:hc}). Here the convention is to leave the strain untouched and instead adjust the PTA sensitivity curve with observation time, again to satisfy (\ref{eq:hc}). This discrepancy is an unfortunate result of the conventions in use by the different GW communities; however it is also natural given the sources under observation. When observing a transient source, such as a burst or inspiral, which changes within the lifetime of the detector it is natural to consider the detector as performing constantly while the signal changes. However when observing a monochromatic source or a stochastic background which is unchanging over the detector lifetime it is more natural to consider the source as being fixed and the sensitivity of the detector gradually improving. All that is required by the definition in (\ref{eq:hc}) is that the ratio $h_{c}(f)/h_{n}(f)$ is constant.

}
